---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# forestFragR

<!-- badges: start -->
<!-- badges: end -->

The goal of forestFragR is to provide simple functions for calculating forest fragmentation and landscape metrics from raster data. The package currently includes functions that wrap functionality from the landscapemetrics package to make forest fragmentation analysis easier and more reproducible.

## Installation

You can install the development version of forestFragR from GitHub with:

``` r
devtools::install_github("PBirungi/forestFragR")
```

## Example

This is a basic example which shows you how to calculate the number of patches (NP) in a raster landscape.

```{r example}
library(forestFragR)
library(terra)

# create a simple raster for demonstration
r <- rast(nrows = 10, ncols = 10)
values(r) <- sample(1:3, 100, replace = TRUE)

# visualize the raster
plot(r, main = "Example Landscape Raster")

# calculate number of patches
calculate_np(r)
```

The function returns a tibble containing the number of patches for each land-cover class in the raster.

