---
title: "Forest Fragmentation Analysis Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Forest Fragmentation Analysis Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Forest fragmentation is a major driver of biodiversity loss,
affecting habitat structure, ecosystem processes, and species
movement. Understanding how forest patches are distributed and
connected across landscapes is essential for ecological analysis
and conservation planning.

The `forestFragR` package provides a workflow for analyzing forest
fragmentation using raster-based land cover data. It includes tools
for preprocessing data, identifying forest patches, computing
landscape metrics, and assessing spatial connectivity.

This vignette demonstrates a complete workflow using both real-world
data and a simplified reproducible example.

---

## Load Packages

```{r setup, message=FALSE, warning=FALSE}
library(terra)
library(forestFragR)
```

## Study Area and Data

This analysis is based on a real-world dataset consisting of a
university forest area of interest (AOI) and land cover data from
the ESA WorldCover dataset.

The AOI defines the spatial boundary of the study, while the land
cover raster provides classification values used to extract forest
areas.

```{r real-data, eval=FALSE}
# Load AOI (vector data)
aoi <- terra::vect("data/university_forest.gpkg")

# Load ESA land cover raster
landcover <- terra::rast("data/esa_landcover.tif")

# Prepare and align datasets
data_prepared <- prepare_data(landcover, aoi)

raster <- data_prepared$raster
aoi <- data_prepared$aoi

```

## Example Data (Reproducible)

To demonstrate the workflow in a reproducible way, we use a small
simulated raster dataset and a simple area of interest (AOI).

```{r example-data}
# Create raster
raster <- rast(nrows = 10, ncols = 10)
values(raster) <- sample(c(10, 20, 30), 100, replace = TRUE)

# Create a simple AOI (extent of raster)
aoi <- as.polygons(ext(raster))

plot(raster, main = "Simulated Land Cover")
plot(aoi, add = TRUE, border = "red")
```

In this example, the value 10 represents forest, while other values
represent non-forest classes.


## Preprocessing Forest Data

The land cover raster is converted into a binary forest/non-forest
map by extracting the forest class.

```{r preprocessing}
forest <- preprocess_forest(raster, forest_class = 10)

plot(forest, main = "Binary Forest Map (1 = forest, 0 = non-forest)")
```

This step simplifies the land cover data and prepares it for
fragmentation analysis.


## Patch Identification

Forest patches are defined as contiguous groups of forest cells.
Identifying these patches is a key step in fragmentation analysis,
as it provides the spatial units for all subsequent landscape metrics.

```{r patches}
patches <- analyze_patches(forest)

plot(patches$patches, main = "Forest Patches")
```

Each patch is assigned a unique identifier and described using
multiple landscape metrics, including area, perimeter, shape complexity and core habitat.

```{r patch-metrics}
head(patches$patch_sizes)
```

These metrics allow for quantitative assessment of fragmentation,
supporting comparisons of patch size distribution, edge effects,
and habitat quality across the landscape.

## Connectivity Analysis

Connectivity between forest patches is assessed using nearest
neighbor distances. This metric quantifies how isolated each patch
is from its closest neighboring patch.

```{r connectivity}
conn <- connectivity_analysis(patches)

conn$nn_distance
```

Lower nearest neighbor distances indicate higher connectivity,
while larger distances suggest greater isolation.


## Visualizing Patch Isolation

Patch isolation is visualized by mapping nearest neighbor distances
onto the raster, allowing spatial interpretation of connectivity.

```{r visualization}
nn_raster <- visualize_patch_isolation(conn)

```

Higher values indicate more isolated patches, while lower values
represent better-connected forest areas.







